---
title: "R Scripts to Model Variance-heterogeneity"
date: "`r Sys.Date()`"
output:
  rmdformats::html_docco:
    highlight: tango
    toc: TRUE
    toc_depth: 4
    fig_width: 12
    fig_height: 8
Author: "Waseem Hussain"
---
  <style type="text/css">
  
  
h1.title {/* Header 1 */
    font-size: 28px;
    font-family:  "Times New Roman", Times, serif;
    color: BLACK;
    #text-transform: Capitalize;
  }
  
h2 { /* Header 2 */
    font-size: 24px;
    font-family: "Times New Roman", Times, serif;
    color: Black;
    text-transform: none;
  }
h3 { /* Header 3 */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: DarkBlue;
    text-transform: none;
  }
  
h4 { /* Header 4 */
    font-size: 16px;
    font-family: "Times New Roman", Times, serif;
    color: Darkred;
    text-transform: none;
  }
  </style>

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               eval=FALSE)
opts_knit$set(width=75)
```

**Load the required libraries**

```{r}
Packages <- c("dglm", "hglm", "gdata","readr","tidyr","ggthemes",
                    "qqman", "CMplot",  "dplyr", "ggplot2", "SNPRelate",
                    "ggcorrplot", "statmod", "pheatmap", "ggfortify")
lapply(Packages, library, character.only = TRUE)
```
**Note: Before showing the codes for DGLM and HGLM analysis, I will first begin with the principal component (PC) analysis of marker data and extract the top fours PCs which is required later for DGLM analysis to account for population structure**

***
## **Principal Component Analysis (PCA)**
***

* In this section we will extract the PC in package SNPRelate using SNP marker data which latter will be used in DGLM analysis to account for population structure.

```{r}
# read the vcf file from the folder
vcf.hww <- "~/Documents/GitHub/vGWAS/Data/Raw_data/genotype.vcf"

# Reformat the data 
genofile<-snpgdsVCF2GDS(vcf.hww, "hww.gds", method="biallelic.only")

#Run PCA
geno_hww<- snpgdsOpen(genofile)
pca_hww1<- snpgdsPCA(geno_hww, num.thread=2)

# variance proportion (%)
pc.percent <- pca_hww1$varprop*100
head(round(pc.percent, 2))

# make a data.frame
pca_data<- data.frame(sample.id = pca_hww1$sample.id,
                      EV1 = pca_hww1$eigenvect[,1],    # the first eigenvector
                      EV2 = pca_hww1$eigenvect[,2],    # the second eigenvector
                      EV3 = pca_hww1$eigenvect[,3],
                      EV4 = pca_hww1$eigenvect[,4],
                      EV5 = pca_hww1$eigenvect[,5],
                      stringsAsFactors = FALSE)
head(pca_data)

#save the pca data as r objective for future use
saveRDS(pca_data, file="~/Documents/GitHub/vGWAS/Data/pca_data/pca_hww_299.rds")
```
***
## **Population Structure of HWW panel**
***

* Here we will draw the population structure of Hard-winter wheat association panel (HWW) given in supplementary materials. 
* First we will add the population information based on type of the grain.

```{r}
# First let us uplaod the information regarding type of wheat
info<-read.csv(file="~/Documents/GitHub/vGWAS/Data/pca_data/information_wheat_type.csv",
               header = TRUE)

# merge with pca data file
pca_data_final<-merge(pca_data, info, by="sample.id")

# Draw the PCA plot 
pca<- pca_data_final[c(2:7)]
autoplot(prcomp(pca[,-6]))
autoplot(prcomp(pca[,-6]), data = pca_data_final, colour = 'Type')+
  xlab("PC1") + ylab("PC2") +
  theme(text=element_text(size=20, family="Times"))+
  geom_hline(yintercept = 0, linetype="dashed", color="black", size=0.5)+
  geom_vline(xintercept = 0, linetype="dashed", color="black", size=0.5)+
  theme_bw()
```

***  
## **Double Generalized Linear Model (DGLM)**
***

* Here we will be presenting the codes to show how to model variance-heterogeneity using DGLM in the context of GWAS. 
* As the analysis is computationally demanding and time consuming we performed this analysis on Holland Computing Center(HCC) server at University of Nebraska, Lincoln. 
* Here we are just providing the snippet of codes and not showing any outputs.
## Read the marker and phenotypic, and principal components

```{r, eval=FALSE}
# First we will be reading the marker data that is in rds formate and assign it to object geno
  geno<-readRDS(file="~/Documents/GitHub/vGWAS/Data/Raw_data/geno.rds")

# Loading map information file
  map <- readRDS("~/Documents/GitHub/vGWAS/Data/Raw_data/map_final.rds")
  map<-map[,c(1:2)]
  
# Read the cadmium data
  pheno_data<- readRDS(file="~/Documents/GitHub/vGWAS/Data/Raw_data/pheno_ok2_final.rds")
  
# Read the principal components data and assign it to object covar
  covar<-readRDS("~/Documents/GitHub/vGWAS/Data/pca_data/pca_hww_299.rds")
```

### **Step 1**: Create a function for DGLM analysis 
**Description of the code:**

* First we created a function so that we can use this function to run the DGLM analysis for all the markers.
* Five arguments are passed to the function including CT (representing column for particular trait in phenotypic data file, here named as phenos), i (representing markers), geno(representing marker data file), and covar (representing PCs of marker data).
* Here we are fitting top four PCs as covariates to account for population structure.
* The basic syntax of DGLM model we are fitting  is as follows:
```
phenotype=marker_effect+ covariates (modelling mean), marker_effect(modeling dispersion/variability)
```
* Finally we are extracting the variables including coefficients (beta), standard error (s.e), P.mean (p-value for mean), P.disp (p-value for dispersion) and assigning them to data frame 'out'
```{r}
    my.pdglm <- function(cT,i,Phenos,geno,covar)
    {
      y <- Phenos[,cT]
      model <- dglm(y ~ geno[,i] + covar[,2] + covar[, 3] + covar[, 4] + covar[, 5],
                    ~ geno[,i], family = gaussian(link = identity))
      P.mean <- summary(model)$coef[2,4] # Extarct p values for mean part
      P.disp <- anova(model)$Adj.P[2] # Extract P values for dispersion part
      s.model <- summary(model$dispersion.fit)
      beta <- s.model$coef[2,1] # Extarct cofficients
      se <- s.model$coef[2,2] # Extract standard errors
      out <- data.frame(Beta=beta, SE = se,P.mean=P.mean, P.disp=P.disp, stringsAsFactors=FALSE) # Save all the extracted variables in data frame out
      return(out)
      
      #print(i)
    }
```
   
### **Step 2**: Use the for-loop to run function
**Description of the code:**

* First we are creating a data frame "TF" with rows equal to number of markers and columns equal to number of variables extracted from the DGLM analysis above.
* Then we are using for-loop to run the above function *my.pdglm()* for all the markers one by one and save the output in TF file.
* We are using function *try()* to continue the loop if error is encountered as some of the markers do not converge.
* Finally we are saving the output as csv file.

```{r, eval=FALSE}
    TF <- matrix(NA,nrow=dim(geno)[2],ncol=4)
    for(i in 1:dim(geno)[2])
    {
      try(
        {
          outm <- my.pdglm(cT=5,i=i,Phenos=pheno_data,geno,covar)
          TF[i,] <- as.numeric(outm)
          print(i)
        }, silent=TRUE)
    }

# Now add map information and select required columns for Manhattan plot
row.names(TF)<-colnames(geno)
colnames(TF) <- c("Beta", "SE", "P.mean", "P.disp")
TF2<-merge(map, TF)
TF2<-data.frame(merge(map,TF[,3:4],by="row.names",all.x=TRUE)) # add map info
colnames(TF2)<-c("marker", "chrom", "pos", "p.mean","p.disp")

# save the output in folder
write.csv(TF,file= "~/Documents/GitHub/vGWAS/Data/vGWAS_output/dglm_cd.csv",sep=',',
              row.names=FALSE)
```
## **Hierarchical Generalized Linear Model (HGLM)**
**Description**:

* Here we will be modeling variance-heterogeneity using HGLM in hglm package. 
* More details on HGLM modelling for genetics data can be found here [The hglm Package (Version 2.0)](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.478.7447&rep=rep1&type=pdf)
* Again HGLM analysis was performed on Holland Computing Center (HCC) server at University of Nebraska, Lincoln. 
* Here we are just providing the snippet of codes and not showing any outputs.

### **Step 1**: Construct the Z matrix for HGLM analysis

#### First we construct G-matrix ( [VanRanden 2008](https://www-ncbi-nlm-nih-gov.libproxy.unl.edu/pubmed/18946147))
```{r, eval=FALSE}
# First upload the marker data
geno<-readRDS(file="~/Documents/GitHub/vGWAS/Data/Raw_data/geno.rds")

# Scale the marker data
Xs<-scale(geno, center = TRUE, scale = TRUE) 

# Construct G matrix
G<-Xs%*%t(Xs)/ncol(Xs)
dim(G)
```
### **Step 1**: Second get Cholesky decomposition of G matrix
```{r, eval=FALSE}
chol.G<-chol(G)
Z0 <- diag(1, nrow = nrow(G), ncol = ncol(G))
Z<-Z0 %*% chol.G
# Z0 is the identity matrix
```
      
### **Step 3**: Create a function for HGLM analysis 
**Description of the code:**

* First we created a function so that we can use this function to run the hglm analysis for all the markers.
* Six arguments are passed to the function including CT (representing column for particular trait in phenotypic data file, here named as phenos), i (representing markers), Phenos (representing phenotypic data file), X (representing marker matrix), Z(representing Z matrix), and X.disp (representing marker matrix for dispersion part).
* Here we are modeling correlated random effects using Z matrix as random effect which is not possible in DGLM.
* The basic syntax of HGLM model we are fitting  is as follows:
```
phenotype=marker_effect (fixed)+ Z (random) (modeling mean), marker_effect(modeling dispersion/variability)
```
* Finally we are extracting the variables including coefficients (beta), standard error (s.e), P.mean (p value for mean), P.disp (p value for dispersion) and assigning them to data frame 'out'
```{r, eval=FALSE}
# Run hglm model for all SNPs using for loop
    	my.hglm <- function(cT,i,Phenos,Z,X,X.disp)
   	 {
      	y <- Phenos[,cT]
      	y2 <<- Phenos[,cT]
      	outm <-hglm(y=y, X= as.matrix(geno[,i]), Z=chol.G2,  X.disp =as.matrix(geno[,i]) , family = gaussian(link = log))
      	estimates_fix<- outm$fixef 
      	SE_Mean<- outm$SeFe 
      	DF<- outm$dfReFe 
      	DP_Mm<-outm$varFix 
      	DP_RM<- outm$varRanef 
      	estimates_rand<- outm$SummVC1[1] 
      	S.E_rand<- outm$SummVC1[2] 
      	out <- data.frame(estimates_fix=estimates_fix, SE_Mean = SE_Mean, DF = DF, 
                        DP_Mm= DP_Mm,  DP_RM= DP_RM, estimates_rand=estimates_rand, S.E_rand=S.E_rand,
                        stringsAsFactors=FALSE)
      	return(out)
    		}

# Aanalysis for cadmium phenotype
    		TF <- matrix(NA,nrow=dim(geno)[2],ncol=7)
    
    		for(i in 1:dim(geno)[2])
    	{
      		try(
       		 {
         	 outm <- my.hglm(cT=5,i=i,Phenos=pheno_data,Z=chol.G2,X=geno,X.disp=geno)
         	 TF[i,] <- as.numeric(outm)
          	print(i)
       		 }, silent = TRUE)
    		}
```

### **Step 4**: Now build function to process the HGLM output and determine p-values

**Description of the code:**

* Here we will determine the p-values for mean and dispersion part.
* Then we will add map information to the output file.
* Add finally select appropriate columns for Manhattan plot
```{r, eval=FALSE}
# read the raw data file
my.hglm1<-function(pheno, geno1, map1){
# add the column names
colnames(pheno)<-c("estimates_fix", "SE_Mean", "DF", 
                       "DP_Mm","DP_RM", "estimates_rand", "S.E_rand")
# add the marker name and position
markernames<-data.frame(colnames(geno1))
# Now combine the output file
pheno <-cbind(markernames, pheno)
# Now estimate the p-values for mean and dispersion part using library dplyr
pheno<-mutate(pheno,
                  p.mean = 2*pt(-abs(estimates_fix/SE_Mean),df=1),
                  p.disp = 2*pt(-abs(estimates_rand/S.E_rand),df=1)
    				)
# now remove the markers with NA values in the file
pheno<- pheno %>% filter(!is.na(p.mean))
# match the markers between map file and outfile
colnames(pheno)<-c("marker", "estimates_fix", "SE_Mean", "DF", "DP_Mm", "DP_RM",
                       "estimates_rand", "S.E_rand", "p.mean","p.disp")
map<-map
# now combine the mapfile and outputfile
    		pheno<-cbind(map, pheno)
# now select the appropriate columns for Manhattan plot
pheno<-select(marker, chrom, pos, p.disp) 
colnames(pheno)<-c("marker", "chrom", "pos", "P")
  		}

#Now use the function above to extract hglm output and save it. 
# read the file obtained from hglm output
hglm_CD<-TF

# load marker data
geno1<-(file="~/Documents/vGWAS/Data/Raw_data.rds")

# now add the map info to the file for Manhattan plots
map_final<-(file="~/Documents/vGWAS/Data/Raw_data/map_final.rds")

# Now use the function
hglm_CD_final<-my.hglm1(pheno= hglm_CD, geno1=geno1, map1 =map_final)

# save the file in folder
write.csv(hglm_CD_final, file="hglm_cd.csv", sep=',', row.names = FALSE)
```

***
## **Manhattan Plot**
***

* Here we will draw the circular Manhattan plot using p-values including P.mean (equivalent to Traditional GWAS),  P.dispersion (obtained in DGLM analysis), and P.dispersion (obtained through HGLM analysis): Figure 2 in the manuscript.

### Load the GWAS, DGLM and HGLM outputs
**Description of the code:**

* Here we will load the data containing p.values for all the three including traditional GWAS, dglm and hglm analysis.
* We will create function to upload all the CSV files
* Combined the p-values from all the three files into one file and rename the columns
```{r}
# First we will create a function to upload all the 3 csv files function
      setwd("~/Documents/GitHub/vGWAS/Data/vGWAS_output")
      import.all<-function(mypath,mypattern,...)
      {
        tmp.list.1<-list.files(mypath, pattern=mypattern)
        tmp.list.2<-list(length=length(tmp.list.1))
        for (i in 1:length(tmp.list.1)){tmp.list.2[[i]]<-read.csv(tmp.list.1[i],...)}
        names(tmp.list.2)<-tmp.list.1
        tmp.list.2
      }
      
# Now upload the all the files
csv.import<-import.all("~/Documents/GitHub/vGWAS/Data/vGWAS_output", "csv$",sep=",")

# here we define the separator of entries in the csv files to be comma.
gwas_cd<-csv.import$gwas_cd.csv
dglm_cd<-csv.import$dglm_cd.csv
hglm_cd<-csv.import$hglm_cd.csv

# subset the data
colnames(gwas_cd)<-c("marker", "chrom", "pos", "gwas_p")
dglm_cd<- dglm_cd[, c(1,5)]
colnames(dglm_cd)<-c("marker","dglm_p.disp")
hglm_cd<-hglm_cd[, c(1,4)]
colnames(hglm_cd)<-c("marker", "hglm_p.disp")

#Now merge all the three data files

all_p<-Reduce(merge, list(gwas_cd, dglm_cd, hglm_cd))
```

### Draw the Circular  Manhattan plot

```{r, eval=FALSE}
setwd("~/Documents/GitHub/vGWAS/Figures")
CMplot(all_p,plot.type="c",r=0.4, col=matrix(c("grey30","slategray", NA,"lightskyblue1","lightseagreen", NA, "thistle","pink", NA),3,3,byrow=T),
         chr.labels=paste("",c("1A", "1B", "1D", "2A", "2B", "2D","3A", "3B", "3D","4A", "4B", "4D","5A", 
    "5B", "5D","6A", "6B", "6D", "7A", "7B", "7D", "UN"),sep=""), cir.legend=TRUE,cir.legend.cex=0.8,
         threshold=1e-5, outward=FALSE,amplify=TRUE,
         threshold.lty=c(1,2), signal.line=1,signal.col="orangered",threshold.col="blue",
         threshold.lwd=0.5, signal.cex=1.5,signal.pch=19, cir.chr.h=1.5,chr.den.col="black",file="jpg",memo="",dpi=300)
```

***
## **Heat Map of Epistatic Interactions**
***

* Here we are providing the sample codes used to draw the heat map of epistasis between markers associated with variance-heterogeneity (Figure 3 of the manuscript).

```{r, eval=FALSE}
# Read the p-value data file
epis_pvalues<-read.csv(file="~/Documents/GitHub/vGWAS/Data/Epistasis_output/epistasis_pvalues.csv",
                       header=TRUE, row.names = 1)

# Convert NA into zeros all the values into -log10
epis_pvalues1<-as.matrix(epis_pvalues)
epis_pvalues1<- -log10(epis_pvalues1)
epis_pvalues1[is.na(epis_pvalues1)]<-0.00

# Draw the heat map
png(file="~/Documents/GitHub/vGWAS/Figures/epistasis_heatmap.png",
    width=12, height =9, units = 'in',res=600)
plot<-pheatmap(epis_pvalues1, cluster_rows = F, cluster_cols = F,scale = "none", legend_labels = "-log10(p-value",fontsize=12, angle_col =90)
plot
dev.off()

```

***
## **Violin plots**
***

* Here we are providing sample codes used to draw the violin plots to see differences in means and variances for grain cadmium concentration (Figure 4 of the manuscript).. 

```{r, eval=FALSE}
# showing for 2B chromsome
data_2b_final<-read.csv(file="~/Documents/GitHub/vGWAS/Data/Raw_data/data_2b.csv",
                        header = TRUE)
ggplot(data_2b_final, aes(x=marker, y=avg1))+
  geom_violin(aes(fill = Genotype), trim = FALSE, position = position_dodge(0.9))+ 
  geom_hline(yintercept = 0.01309, color = "darkred", size =0.6,show.legend = TRUE,linetype = 2)+
  geom_boxplot(aes(fill = Genotype), width = 0.2, position = position_dodge(0.9))+
  scale_fill_manual(values = c("#00AFBB", "#E7B800"))+
  labs(title = "", y="Cadmium conc. (mg/kg)", x="Markers")+
  theme_few() + #change background of the plot
  theme(plot.title = element_text(color="black",size=12, hjust=0.5))+
  theme(axis.text.x=element_text(colour='black', size=12)) + #asthetics of x-axis text
  theme(axis.text.y=element_text(colour='black', size=12)) + #asthetics of y-axis text
  theme(axis.title.x = element_text(colour='black', size=12, vjust=0.0)) + #asthetics of x-axis title
  theme(axis.title.y = element_text(colour='black', size=12))+
  theme(legend.title = element_text(colour="black", size=14),
        legend.text = element_text(colour="grey0", size=11))+
  guides(fill=guide_legend(title="Genotype"))
```
##### ===================END===================