---
title: "Correction for Multiple testing"
output: github_document
---

<style type="text/css">
  
  
h1.title {/* Header 1 */
    font-size: 28px;
    font-family:  "Times New Roman", Times, serif;
    color: BLACK;
    #text-transform: Capitalize;
  }
  
h2 { /* Header 2 */
    font-size: 24px;
    font-family: "Times New Roman", Times, serif;
    color: Black;
    text-transform: none;
  }
h3 { /* Header 3 */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: DarkBlue;
    text-transform: none;
  }
  
h4 { /* Header 4 */
    font-size: 16px;
    font-family: "Times New Roman", Times, serif;
    color: Darkred;
    text-transform: none;
  }
  </style>

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               eval=FALSE)
opts_knit$set(width=75)
```

**Load the required libraries**
```{r}
      Packages <- c("dglm", "hglm", "gdata","readR","tidyr","ggplot",
                    "qqman", "CMplot",  "dplyr", "ggplot2", "SNPRelate",
                    "ggcorrplot", "statmod")
    lapply(Packages, library, character.only = TRUE)

```
## **Multiple testing**

* We will be using [Li and Ji (2005)](https://www.nature.com/articles/6800717) method for multiple testing and choose significance threshold level.
The reason we use [Li an Ji (2005)](https://www.nature.com/articles/6800717)is to make sure to account for correlated markers (LD between markers) and determine independent number of tests.

Below is the code for multiple testing:

### Effective number of tests(Meff)
```{r}
# Read the orginal marker data used for DGLM and HGLM analysis
      setwd("~/Documents/GitHub/vGWAS/Data/Raw_data")
      geno<-readRDS(file="geno.rds")
      geno<-as.matrix(geno)
# get correlation matrix
      corr.matrix <- cor(geno)
# get effective number of tests
# MEFF
      meff<- meff(corr.matrix, method = "liji")
### Determine threshold significance level
      p_threshold= (1- (1-0.05))^1/meff
      p_threshold
```
